Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-09-04 response.tex

\response{We have added a comparison of the computing time required for each method in Section 4 of the revised manuscript. We have also added a comparison of edgeR estimates to eBayes (Laplace) in Figure 3 which show shrinkage for half-parental differences and hybrid effects.}

\response{Added in Section 4. The bottom line is edgeR takes seconds, Ji et al. takes minutes, the other methods take hours.}

\comment{Page 1.  I was instructed that heterosis is merely characterized by the hybrid performance not being equal to the average of the two parental lines.  So in your very first sentence, I might write “Heterosis, or hybrid vigor, occurs when hybrid progeny display phenotypes that are superior to the average phenotypes of their parents.” i.e., I might be only interested in the $\delta_g$’s. However, your later characterization of heterosis is really something called overdominance to me…where the hybrid performance exceeds the performance of either parent.  I wonder if you could then be more specific in your definitions in your first sentence…and also in your third sentence in this particular paragraph. lines 48-59.  Perhaps this was already captured well in the Ji et al. (2014) paper but why not simply test heterosis as the difference between the hybrid and the average parental performance (i.e. $\mu_3 - 0.5 \mu_1 - 0.5 \mu_2$).  Again, I think your issue pertains really to an overdominance test of some sort.}

\response{Many publications mentioning heterosis do not define it clearly.  For example, the 2013 {\em Nature Reviews Genetics} paper by Chen that we now cite in our revised manuscript states the following definition: ``Heterosis (Also known as hybrid vigour).  When hybrids display increased levels of growth, survival or fitness relative to their parents.''  Like Chen, some other author present heterosis as synonymous with hybrid vigor, and then go on to describe hybrid vigor as the phenomenon in which the hybrid outperforms the parents.  It may not be so clear what ``outperforms the parents'' means, but the idea behind hybrid vigor and the reason it is of such interest (particularly in maize) and so useful is that the hybrid is better than {\bf both parents}, not simply better than the average.  That said, we do agree with your definition of heterosis.  We originally wrote that first sentence so that it would be true regardless of which definition of heterosis one takes.  We were saying that heterosis occurs when hybrid vigor occurs.  The reverse is not necessarily so, and we did not claim it to be in that first sentence.  However, we acknowledge that our first sentence was misleading because most readers would naturally take the sentence for a definition.  Furthermore, for simplicity, we did start using ``heterosis'' as a synonym for the situation where the hybrid mean expression is more extreme than both parents.  That is really what we are interested in because the phenotypic traits we see in maize have this pattern, so it is only natural to seek genes whose expression has this same pattern.  In the revised manuscript, we have been more careful with terminology.  We have defined heterosis to be a difference between the hybrid mean and the average of the parental means.  We have also introduced a new term (extreme heterosis, EH) to specify the situation where the hybrid mean is more extreme than both parental means.  We continue to use low-parent heterosis (LPH) and high-parent heterosis (HPH) to distinguish the two types of EH.  We also explain in the manuscript why we prefer the LPH/HPH terminology to underdominance/overdominance.  In our experience, these latter terms are used to describe the ``gene action'' at a single locus with respect to some quantitative trait.  We can't asses the action of any locus on any of our expression traits in data from inbred and hybrid crosses.}

@@

\comment{Page 7, line 15.  Why delete the low counts?  Data quality issues?}

\response{ShrinkBayes failed for us when too many zeros were present. We do not believe the comparison of other methods would change if we did not remove genes with low counts.  In practice, it is common to remove low count genes from RNA-seq datasets prior to analysis because these genes contain little information about differential expression and lead to computational difficulties for some methods. With that being said, we did not remove any genes from the real data analysis.}

\comment{Page 7, lines 17-22.  Can you provide us any information on how well the hyperparameters were estimated based on your empirical Bayes strategy?}

\response{Since we did not simulate from the hierarchical portion of the model, there are no true values of the hyperparameters to compare against. We tried to get at this question by calculating means and standard deviations of the true values for the gene-specific parameters in each simulation. Table \ref{t:hyperparameter} shows root mean squared error (RMSE) of the estimated hyperparameters. Because each simulation has a different value of the hyperparameters (due to different genes randomly being included), the RMSEs are hard to interpret. For this reason, we did not include this table in the manuscript because we did not believe it added much value. }

\input{include/rmse}

\newpage
\comment{Page 8, line 37.  It is not obvious to me why $H_g2*$ (are the two parents the same?) is a necessary component of a heterosis (overdominance) test!?}

@@

\comment{Page 11 line 39.  Is it really “independence” or nearly non-identifiability that is the issue?}

\response{We're sorry, but we don't understand this comment. We believe all the parameters in the model are identified and therefore are not sure what non-identifiability is being referenced.}

\comment{Page 11 last paragraph.  So this then begs an assessment of how good/poor these hyperparameter estimates were.}

\response{See discussion above and Table \ref{t:hyperparameter}.}

### 2015-09-04 ebayes_rnaseq_heterosis.tex
response done

Following \cite{ji2014estimation}, we reparameterize the gene-variety mean structure into the gene-specific parental average ($\phi_g$), half-parental difference ($\alpha_g$), and hybrid effect ($\delta_g$). For our heterosis study where $V=3$, we let $v=1,2$ indicate the two parental varieties and $v=3$ indicate the hybrid. The reparameterization is  
\[ \phi_g = \frac{\mu_{g1}+\mu_{g2}}{2}, \quad 
\alpha_g = \frac{\mu_{g1}-\mu_{g2}}{2}, \quad \mbox{and} \quad
\delta_g = \phi_g-\mu_{g3}. \]

@@

\subsection{Hyperparameter estimation}

As described in Section \ref{s:simulation}, we simulated data only from the negative-binomial portion of the model, but not from the hierarchical distributions. Thus, there are no true values for the hyperparameters to compare against. But, we can compute means and standard deviations of the gene-specific parameters that were used in the simulation study and compare these to the estimated hyperparameter values. 
\begin{table}
\caption{Empirical values calculated from the gene-specific parameters used in the simulation and mean of estimated values across the 10 simulations within replicate number. Maximum standard deviation in the estimates across the 10 simulations was 0.004.}
\label{t:hyperparameter}
\begin{tabular}{l|c|ccc|}
& \multicolumn{1}{c|}{Empirical} & \multicolumn{3}{c}{Number of replicates}\\
Parameter & value & 4 & 8 & 16 \\
\hline
% with (SD)
% $\eta_\phi$ & 4.675 & 4.681 (0.004) & 4.741 (0.002) & 4.817 (0.003) \\
% $\eta_\alpha$ & -0.017 & -0.023 (0.002) & -0.018 (0.003) & -0.014 (0.002)  \\
% $\eta_\delta$ & -0.046 & -0.046 (0.002) & -0.031 (0.002) & -0.022 (0.003)  \\
% $\eta_\phi$ & -2.713 & -2.290 (0.004) &  -2.335 (0.003) & -2.387 (0.002)\\
% \hline
% $\sigma_\phi$ & 1.815 & 1.821 (0.002) & 1.779 (0.002) & 1.726 (0.003)\\
% $\sigma_\alpha$ & 0.364 & 0.459 (0.002) & 0.399 (0.002) & 0.357 (0.002) \\
% $\sigma_\delta$ & 0.145 & 0.389 (0.002) & 0.317 (0.002) & 0.268 (0.001) \\
% $\sigma_\psi$ & 0.737 & 0.401 (0.001) & 0.493 (0.001) & 0.542 (0.001) \\
% 3 decimal places
% $\eta_\phi$ & 4.675 & 4.681 & 4.741  & 4.817  \\
% $\eta_\alpha$ & -0.017 & -0.023  & -0.018 & -0.014  \\
% $\eta_\delta$ & -0.046 & -0.046 & -0.031  & -0.022   \\
% $\eta_\phi$ & -2.713 & -2.290  &  -2.335  & -2.387\\
% \hline
% $\sigma_\phi$ & 1.815 & 1.821  & 1.779  & 1.726 \\
% $\sigma_\alpha$ & 0.364 & 0.459  & 0.399  & 0.357  \\
% $\sigma_\delta$ & 0.145 & 0.389  & 0.317  & 0.268  \\
% $\sigma_\psi$ & 0.737 & 0.401 & 0.493 & 0.542 \\
$\eta_\phi$ & 4.7\phantom{00} & 4.7\phantom{00} & 4.7\phantom{00}  & 4.8\phantom{00}  \\
$\eta_\alpha$ & -0.017 & -0.023  & -0.018 & -0.014  \\
$\eta_\delta$ & -0.046 & -0.046 & -0.031  & -0.022   \\
$\eta_\psi$ & -2.7\phantom{00} & -2.3\phantom{00}  &  -2.3\phantom{00}  & -2.4\phantom{00} \\
\hline
$\sigma_\phi$ & 1.8\phantom{00} & 1.8\phantom{00}  & 1.8\phantom{00}  & 1.7\phantom{00} \\
$\sigma_\alpha$ & 0.36\phantom{0} & 0.46\phantom{0}  & 0.40\phantom{0}  & 0.36\phantom{0}  \\
$\sigma_\delta$ & 0.15\phantom{0} & 0.39\phantom{0}  & 0.32\phantom{0}  & 0.27\phantom{0}  \\
$\sigma_\psi$ & 0.74\phantom{0} & 0.40\phantom{0} & 0.49\phantom{0} & 0.54\phantom{0} \\
\hline
\end{tabular}
\end{table}
Table \ref{t:hyperparameter} provides the computed means and standard deviations from the gene-specific parameter values, called empirical values, and compares these values against the mean of estimated hyperparameters across the 10 simulations within replicate number. For the mean parameters, there is reasonable agreement amongst the different numbers of replicates with a somewhat high estimate for the mean of the overdispersion parameter $(\eta_\psi)$ and, perhaps, a trend for the mean of the hybrid effect ($\eta_\delta$), although we would not have expected this. For the standard deviation parameters, $\sigma_\phi$ is consistent across the replicate numbers, $\sigma_\alpha$ and $\sigma_\delta$ mean estimates are decreasing toward the empirical values as the number of replicates increases, and $\sigma_\phi$ is increasing toward the empirical value as the number of replicates increases. %We expect this pattern for $\sigma_\alpha$ and $\sigma_\delta$ since these are based on moment matching of independent gene-specific parameter estimates. For low numbers of replicates, these independent estimates will have relatively large sampling variability that will cause a bias toward larger values for the moment-matched standard deviation estimates.

>>>>

Following \cite{ji2014estimation}, we reparameterize the gene-variety mean structure into the gene-specific parental average ($\phi_g$), half-parental difference ($\alpha_g$), and hybrid effect ($\delta_g$). For our heterosis study where $V=3$, we let $v=1,2$ indicate the two parental varieties and $v=3$ indicate the hybrid. The reparameterization is  
\[ \phi_g = \frac{\mu_{g1}+\mu_{g2}}{2}, \quad 
\alpha_g = \frac{\mu_{g1}-\mu_{g2}}{2}, \quad \mbox{and} \quad
\delta_g = \mu_{g3}-\phi_g. \]

@@

% \subsection{Hyperparameter estimation}
% 
% As described in Section \ref{s:simulation}, we simulated data only from the negative-binomial portion of the model, but not from the hierarchical distributions. Thus, there are no true values for the hyperparameters to compare against. But, we can compute means and standard deviations of the gene-specific parameters that were used in the simulation study and compare these to the estimated hyperparameter values. 
% \include{include/rmse}
% \include{include/rae}
% % \begin{table}
% % \caption{Empirical values calculated from the gene-specific parameters used in the simulation and mean of estimated values across the 10 simulations within replicate number. Maximum standard deviation in the estimates across the 10 simulations was 0.004.}
% % \label{t:hyperparameter}
% % \begin{tabular}{c|c|ccc|}
% % & \multicolumn{1}{c|}{Empirical} & \multicolumn{3}{c}{Number of replicates}\\
% % Parameter & value & 4 & 8 & 16 \\
% % \hline
% % $\eta_\phi$ & \phantom{-}4.6\phantom{00} & \phantom{-}4.7\phantom{00} & \phantom{-}4.7\phantom{00}  & \phantom{-}4.8\phantom{00}  \\
% % $\eta_\alpha$ & -0.028 & -0.023  & -0.018 & -0.014  \\
% % $\eta_\delta$ & \phantom{-}0.070 & \phantom{-}0.046 & \phantom{-}0.031  & \phantom{-}0.022   \\
% % $\eta_\psi$ & -2.4\phantom{00} & -2.3\phantom{00}  &  -2.3\phantom{00}  & -2.4\phantom{00} \\
% % \hline
% % $\sigma_\phi$ & \phantom{-}1.8\phantom{00} & \phantom{-}1.8\phantom{00}  & \phantom{-}1.8\phantom{00}  & \phantom{-}1.7\phantom{00} \\
% % $\sigma_\alpha$ & \phantom{-}0.45\phantom{0} & \phantom{-}0.46\phantom{0}  & \phantom{-}0.40\phantom{0}  & \phantom{-}0.36\phantom{0}  \\
% % $\sigma_\delta$ & \phantom{-}0.31\phantom{0} & \phantom{-}0.39\phantom{0}  & \phantom{-}0.32\phantom{0}  & \phantom{-}0.27\phantom{0}  \\
% % $\sigma_\psi$ & \phantom{-}0.59\phantom{0} & \phantom{-}0.40\phantom{0} & \phantom{-}0.49\phantom{0} & \phantom{-}0.54\phantom{0} \\
% % \hline
% % \end{tabular}
% % \end{table}
% Table \ref{t:hyperparameter} provides the computed means and standard deviations from the gene-specific parameter values, called empirical values, and compares these values against the mean of estimated hyperparameters across the 10 simulations within replicate number. For the mean parameters, there is reasonable agreement amongst the different numbers of replicates with a somewhat high estimate for the mean of the overdispersion parameter $(\eta_\psi)$ and, perhaps, a trend for the mean of the hybrid effect ($\eta_\delta$), although we would not have expected this. For the standard deviation parameters, $\sigma_\phi$ is consistent across the replicate numbers, $\sigma_\alpha$ and $\sigma_\delta$ mean estimates are decreasing toward the empirical values as the number of replicates increases, and $\sigma_\phi$ is increasing toward the empirical value as the number of replicates increases. %We expect this pattern for $\sigma_\alpha$ and $\sigma_\delta$ since these are based on moment matching of independent gene-specific parameter estimates. For low numbers of replicates, these independent estimates will have relatively large sampling variability that will cause a bias toward larger values for the moment-matched standard deviation estimates.
